---
title: 'Data cleaning: easy scraping, reformatting and reshaping'
author: "Liz Lucas, IRE & Yanqi Xu & Jasmine Ye Han"
output: html_document
---
```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning=FALSE)
```

## Intro 

In this session we're going to:

**1. Get data into R!**
We will scrape an html table and load data from Excel, csv and csv files on the web.

**2. Clean data in R!**
We will practice cleaning up column names, converting data types, combining multiple data sets, manipulating strings. There's also some reshaping at the end because it's crucial step to get data ready for analysis.

You will walk away with: 
- basic understanding of loading and cleaning data in R
- a list of helpful packages and functions
- some resources for further learning 

## Let's go!

### Load packages

We'll need to load several packages for this class.  

```{r}
library(tidyverse) # we'll be using mostly tidyr and stringr packages within the tidyverse, handy for data cleaning and string manipulation respectively.
library(lubridate) # this package allows us to manipulate dates
library(here) # easier file paths
library(readxl) # read and write data from Excel files
library(googlesheets4) #write to and pull data from Google Sheets
library(janitor) #data cleaning package - clean_names() is great
library(campfin) # developed by the Accountability Project to clean campaign finance data but has tons of tools for data cleaning
library(rvest) #web scraping
```


### Loading/scraping an html table

Let's say an EPA webpage has a [table](https://www.epa.gov/nutrientpollution/estimated-nitrate-concentrations-groundwater-used-drinking) with some useful information: the percentage of state area with groundwater nitrate contamination for all states. You want to know how Nebraska is doing compared to other states, but there's no sort button to view a column in descending order. So let's scrape it.

Rvest package can help us do that: 
- `read_html()` function: 
  - performs a HTTP request then parses the HTML received with xml2 package, a lower level package that Rvest buils on top of.
  - returns a xml_document object which you’ll then manipulate using rvest functions. xml_document is a class from xml2 package.
  
- `html_element()` function: 
  - finds HTML element using CSS selectors or XPath expressions
    
- `html_table()` function: 
  - parses an html table into a data frame
  
Tips:
- Right click the page and "inspect" allows you to peak into the HTML tree and find what elements you want. A more detailed tutorial on how to do that  [here](https://rpubs.com/Jasmineyehan/582039)
- Learn CSS selectors with an interactive game:  https://flukeout.github.io/
  
```{r}
epa_url <- "https://www.epa.gov/nutrient-policy-data/estimated-nitrate-concentrations-groundwater-used-drinking"
## creating a variable for the url

epa_tab <- read_html(epa_url) %>% ##read in the html page
  html_element("table") %>% ##select the "table" element
  html_table() ##parse the html table into data frame

# Clean up column names
names(epa_tab) <- c("state","area","percent","private_perc_2005","private_perc_2015")
```

Now you can manipulate the data frame. But first let's check the data types with `str()`.

```{r}
str(epa_tab)
```
Some columns are characteristics, but to perform any sorting they need to be numeric. Let's use the following functions to convert them:

- `mutate()`: 
  - let you create new columns that are functions of existing variables
  - if you just want to create one new column, i.e. based on "area" column: epa_tab <- epa_tab %>% mutate(area = as.numeric(area))
  
- `across()`: 
  - makes it easy to apply the same transformation to multiple columns

```{r}
## to double check: epa_tab %>% mutate(across(2:4, as.numeric)) %>% View()
epa_tab <- epa_tab %>% mutate(across(2:4, as.numeric))
```
You might have seen "Warning: NAs introduced by coercion"; it's because when it says "no data" they are character strings that R couldn't turn into numeric. R coerced them to be NAs.

Now let's rank the states based on the percentage of area with nitrate contamination, from highest to lowest.

```{r}
epa_tab %>% arrange(desc(percent))
```

Now that the data are numeric you can also perform calcuations of multiple columns, for example, the change in % of residents on a private well in 2015 vs 2005.
 
```{r}
epa_tab <- epa_tab %>% 
  mutate(perc_change = private_perc_2015-private_perc_2005)
```

### Data from Excel

We are going to work with court records from an eviction court in Nebraska. The data was maintained by the administrative office of the courts, provided to the Flatwater Free Press by a researcher who requested it in the first place.

```{r}
eviction_p1 <- read_xlsx("../data/evictions 2020-01-01 to 2021-01-01.xlsx") %>% clean_names()

eviction_p2 <- read_xlsx("../data/evictions 2019-04-01 to 2019-12-31.xlsx") %>% clean_names()

eviction_p3 <- read_xlsx("../data/evictions 04-01-2017 to 03-31-2019.xlsx") %>% clean_names()
```

#### Combining multiple datasets into one

We can easily combine data files in the same structure with `bind_rows()`. Note that these data column types need to be consistent, so we can first make sure the date columns are read in as dates. 

We can look at the specific columns using `select()`. And if we don't want to type the column names for each, we can use `ends_with()` or `start_with()` to select those columns based on string pattern.

```{r}
eviction_p1 %>% select(ends_with("date"))
eviction_p2 %>% select(ends_with("date"))
eviction_p3 %>% select(ends_with("date"))
## read more about POSIX classes here:https://www.stat.berkeley.edu/~s133/dates.html
```
Same with the zipcode columns.
```{r}
eviction_p1 %>% select(starts_with("zip"))
eviction_p2 %>% select(starts_with("zip"))
eviction_p3 %>% select(starts_with("zip"))
## read more about POSIX classes here:https://www.stat.berkeley.edu/~s133/dates.html
```

#### Converting dates and numbers

[lubridate package](https://github.com/ireapps/R-for-data-analysis-2022/blob/main/docs/lubridate.pdf) has some handy functions for converting dates: such as `ymd()`, which parses dates in the year month date order. P.S. as long as formatting goes it works for all of these: "yyyy-mm-dd",""yyyy/mm/dd","yyyymmdd".

```{r}
eviction_p1 <- eviction_p1 %>%  
  mutate(across(.cols = ends_with("date"), ymd)) %>% 
  mutate(across(.cols = starts_with("zip"), as.double))

eviction_p2 <- eviction_p2 %>%  
  mutate(across(.cols = ends_with("date"), ymd)) %>% 
  mutate(across(.cols = starts_with("zip"), as.double))
  
eviction_p3 <- eviction_p3 %>%  
  mutate(across(.cols = ends_with("date"), mdy)) %>% 
  mutate(across(.cols = starts_with("zip"), as.double))

```

we need to use `mdy()` for eviction_p3 because the date columns have a different format from the other two: "mm/dd/yyyy" rather than "yyyy-mm-dd".

#### Binding rows

Now that we've got a working date field, we can combine the data and save to a new variable.

```{r}
eviction <- eviction_p1 %>% bind_rows(eviction_p2) %>% bind_rows(eviction_p3) 
```

#### extracting year out of date
We can extract the year from the `file_date` field with the `year()` function in lubridate.

```{r}
eviction <- eviction %>% mutate(file_year = year(file_date))
```

#### String functions

Now the fun begins. We need to isolate plaintiff in the case name and which ones are tied to Omaha Public Authority. 

##### Split up a string
- `str_extract()`: extracts the first complete match from each string.

You put [regular expressions](https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html) in the function as an argument to tell R what pattern to match. 

- Find regex cheat sheet [here](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf) 
- Test your regex [here](https://regex101.com/)

Look for these on the cheat sheet and try running the example code:
- Match characters:
  - `\\.`
  - `\\s`
  - `.` 
- Anchors: 
  - `^a`
  - `a$`
- Look-arounds:
  - `a(?=c)`
  - `(?<=b)a`
- Quantifiers:
  - `a+`

```{r}
eviction <- eviction %>% 
  mutate(caption_up = str_to_upper(eviction$caption), .after = "caption") 
## create a "caption_up" column after "caption" column, which is everything in "caption" column to upper case 

eviction <- eviction %>% 
  mutate(
    plaintiff = caption_up %>% str_extract("^.+(?=\\sV\\.)") %>% str_trim(), 
    def = caption_up %>% str_extract("(?<=\\sV\\.).+$") %>% str_trim()
  )
## plaintiff: look for start of string, what's followed by ' V.'
## defendant: look for end of string, what's proceded by V.

landlord_names <- eviction %>% tabyl(plaintiff)
```

#### Determine if string matches a pattern

We have 5,000+ distinct landlord names. Now we can use a `str_detect` function to capture the ones denoting the Omaha Housing Authority. We don't know what these are, so let's cast a wide net first. 

But let's first narrow the data to everything in Douglas County. We can use the `str_detect` function to tell R to give us all the entries whose `court_name` variable contains the string "Douglas County".

```{r}
eviction <- eviction %>% 
  filter(court_name %>% str_detect("Douglas County"))
```

Next we will look for every possible name variation of Omaha Housing Authority and generate a frequency table
```{r}
oha <- eviction %>% 
  mutate(caption = str_to_upper(caption)) %>% 
  filter(caption %>% str_detect(str_to_upper("Omaha Housing Authority|OHA|Housing Authority|Housing Authority-City of Omaha|Housing Auth. City of Omaha|Housing Authority Of Omaha|Omaha Housing Auth|Housing Auth. City of Omaha|Housing Auhtority City of Omaha|Housing Authority City Omaha|^OMA")))

## | is an alternator, means "OR", allows for searching multiple phrases;
## ^ is an anchor, means "starts with"


oha_names <- oha %>% tabyl(plaintiff) 
## janitor::tabyl() generates a frequency table, Returns a data.frame with frequencies and percentages of the tabulated variable
```

We can manually review the plaintiff names and see if they're truly OHA. 
Let's check if "THE HOUSING AUTHORITY" is truly OHA cases. By checking the address, we can manually verify these cases. 

When you need to quickly inspect a certain column, use `select` to rearrange column positions for easier viewing.

```{r}
oha %>% filter(plaintiff == "THE HOUSING AUTHORITY") %>% 
  #addr1 now shifts to the left and everything else stays in place
  select(addr1, everything())
```

Once we find the plaintiff names that are actually OHA, let's save them to an object.
```{r}
oha_names$plaintiff[c(7:13,30:34,59:67)] -> oha_clean_names
```

We will filter the plaintiff column based on if the plaintiff name matches any of names in oha_clean_names.

```{r}
#cases we are confident that are filed by OHA 
actual_oha <- oha %>% filter(plaintiff %in% oha_clean_names)
# the ones that are not OHA
oha_out <- eviction %>% filter(plaintiff %out% oha_names$plaintiff)

oha_out %>% tabyl(plaintiff) -> oha_out_names
```

Now we have a clean data set of all OHA filings, let's see how many cases they filed each year, and the ratio of OHA filings out of all filings.

`count()` in dplyr package lets you quickly count the unique values of one or more variables.

```{r}
# OHA cases each year
x <- actual_oha %>% count(file_year)
names(x) <- c("filing_year","oha_cases")

# Let's see how many cases in total are filed each year
y <- eviction %>% count(file_year)
names(y) <- c("filing_year","all_cases")

# join and calculate percent
z <- x %>% left_join(y) %>% 
  mutate(oha_perc = oha_cases/all_cases)

z %>% head()
```


### Working with Google Sheets

Say you'd like to send this to another reporter in your newsroom for some spot checks to see if you've gotten everything right. An easy way to do this is connecting to your Google account and upload the sheet to a Google Drive. 

Let's get started. First, you'll need a Google Account and install the `googlesheets4` package. 

Let's create a sheet from scratch. You'll first need to authenticate your account.

```{r}
googlesheets4::gs4_auth()
```

Next we will create a Googlesheet 
```{r}
googlesheets4::gs4_create("oha",sheets = actual_oha)
# first argument is the sheet name, second argument is the data frame, which will also be the sheet name
```

You can then make changes in the Google sheet. Grab the ID of your sheet. It's in the address url. The string after "https://docs.google.com/spreadsheets/d/" 

Let's add the summary table to the same Google sheet in another tab.

```{r}
sheet_write(z, ss = "1yyMApNbIVAWxvFPEw_11ZrkRQdip9to_UIs7GL4Op74",sheet = "smmary_table")
```

### Data reformatting & reshaping

#### Adding description columns

Next we'll take a look at `osha`.

```{r}
osha <- read_csv("../data/osha.csv")

# or read from github repo
# osha <- read_csv("https://raw.githubusercontent.com/ireapps/teaching-guide-R123/main/data/osha.csv")
```

```{r}
glimpse(osha)
```

If you look through the [documentation](https://www.osha.gov/sites/default/files/ITA_Data_Dictionary.pdf) for this dataset, you'll notice that some of these fields are coded, such as `size` and `establishment_type`. For columns that have many value options, we might want to join to a lookup table. But for just a few values, we can add a `_desc` column into our data and code in values based on the original column.

We'll add an `estab_type_desc` column based on the values in `establishment_type`, using a function called `case_when()`. This is something like an `if` or `ifelse` statement:

```{r}
# test it out
osha %>% mutate(estab_type_desc = case_when(
  establishment_type==1 ~ "Not a Government Entity",
  establishment_type==2 ~ "State Government Entity",
  establishment_type==3 ~ "Local Government Entity",
  TRUE ~ "Error"
)) %>% 
  count(establishment_type, estab_type_desc)
```

```{r}
# make it permanent
osha <- osha %>% mutate(estab_type_desc = case_when(
  establishment_type==1 ~ "Not a Government Entity",
  establishment_type==2 ~ "State Government Entity",
  establishment_type==3 ~ "Local Government Entity",
  TRUE ~ "Error"
))
```

### Reshaping data structure

This is the original file that from the Census Bureau. 

```{r}
poverty <- read_csv("../data/poverty_original.csv")

#poverty <- read_csv("https://raw.githubusercontent.com/ireapps/teaching-guide-R123/main/data/poverty_original.csv")
```

In this example, each variable (i.e. `below50`, `below125`, etc) is its own row. To make it easier to do calculations by county, I transposed this data so that each variable would be its own column rather than row. I did that using `pivot_wider()` (for the sake of this example, I'm going to exclude the margin of error, or `moe`).

```{r}
poverty %>% 
  select(-moe) %>% 
  pivot_wider(names_from=variable, values_from=estimate)
```

There is a function called `pivot_longer()` that does the opposite:

```{r}
# First I'll create a new variable with the wider data:
poverty_wide <- poverty %>% 
  select(-moe) %>% 
  pivot_wider(names_from=variable, values_from=estimate)

# Then I'll turn it back to long using pivot_longer()
poverty_wide %>% pivot_longer(cols = population:below500, names_to="variable", values_to="estimate")
```
